---
title: "Metadata, Schema, and Backup in airtable2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metadata, Schema, and Backup in airtable2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

```{r setup}
library(airtable2)
```

## Overview

airtable2 provides two related but distinct capabilities for working with the
structure of your Airtable bases:

1. **Reading and writing human-readable descriptions** — the metadata attached
   to tables and fields. These are the descriptions you see when you hover over
   a field in the Airtable UI, and they can be managed programmatically with
   `air_schema()`, `air_meta()`, `air_meta_push()`, and `air_meta_sync()`.

2. **Full structural and data backup/restore** — a complete snapshot of a base
   including its schema and all record data, managed with `air_dump()` and
   `air_restore()`.

Both capabilities serve the principle of *documentation as code*: your base's
structure and descriptions should live in version-controlled files alongside
the R code that uses them. They also support disaster recovery (restoring a
base after accidental deletion) and base migration (moving data between
workspaces).

See the [Getting Started](airtable2.html) vignette for authentication setup
and basic read/write operations. For field types that require special handling
(multi-select, attachments, linked records), see the
[Special Types](data-types.html) vignette.

---

## Schema inspection with `air_schema()`

`air_schema()` returns a tidy tibble describing every table and field in a
base. It is the fastest way to understand the structure of a base before
reading data or writing code against it.

```{r air-schema-call}
schema <- air_schema("appXXXXXX")
schema
```

The result has one row per table, with a `fields` list-column containing a
tibble for each table's fields:

```
# A tibble: 3 x 4
  table_id        table_name   table_description fields
  <chr>           <chr>        <chr>             <list>
1 tblABCDEFGHIJKL Projects     Project registry  <tibble [8 x 4]>
2 tblMNOPQRSTUVWX People       Team members      <tibble [5 x 4]>
3 tblYZABCDEFGHIJ Status Codes Lookup values     <tibble [3 x 4]>
```

Drill into a single table's fields:

```{r air-schema-fields}
schema$fields[[1]]
```

```
# A tibble: 8 x 4
  id              name          type                  description
  <chr>           <chr>         <chr>                 <chr>
1 fldAAAAAAAAAAAAAA Name         singleLineText        Primary project name
2 fldBBBBBBBBBBBBBB Status       singleSelect          Current project phase
3 fldCCCCCCCCCCCCCC Owner        multipleRecordLinks   Link to People table
4 fldDDDDDDDDDDDDDD Budget       currency              Approved budget (USD)
5 fldEEEEEEEEEEEEEE Start Date   date                  NA
6 fldFFFFFFFFFFFFFF End Date     date                  NA
7 fldGGGGGGGGGGGGGG Progress     formula               Computed from milestones
8 fldHHHHHHHHHHHHHH Record ID    autoNumber            NA
```

**Common use cases for `air_schema()`:**

- **Auditing field types** before writing — confirm that a `currency` field
  will be read as `numeric`, that `multipleSelects` will arrive as a
  list-column, and so on. The [Special Types](data-types.html) vignette maps
  Airtable types to their R representations.
- **Programmatic base introspection** — extract field names for a specific
  table to build dynamic column selections or formulas.
- **Checking for computed fields** before a restore — fields with types like
  `formula`, `rollup`, `autoNumber`, and `lookup` cannot be written via the
  API and will need to be re-created manually.

```{r air-schema-computed}
# Find computed fields that cannot be restored
fields_df <- schema$fields[[1]]
computed_types <- c(
  "formula", "rollup", "lookup", "count",
  "autoNumber", "createdTime", "lastModifiedTime",
  "createdBy", "lastModifiedBy", "externalSyncSource", "aiText", "button"
)
fields_df[fields_df$type %in% computed_types, c("name", "type")]
```

---

## Self-documenting bases with `air_meta()`

Where `air_schema()` returns a nested tibble organised by table, `air_meta()`
returns a *flat* tibble — one row per field across all tables — including the
human-readable `description` stored in Airtable's metadata API.

```{r air-meta-call}
meta <- air_meta("appXXXXXX")
meta
```

```
# A tibble: 16 x 6
   table_name   table_id        field_name   field_id         field_type          description
   <chr>        <chr>           <chr>        <chr>            <chr>               <chr>
 1 Projects     tblABCDEFGHIJKL Name         fldAAAAAAAAAAAAA singleLineText      Primary project name
 2 Projects     tblABCDEFGHIJKL Status       fldBBBBBBBBBBBBBB singleSelect       Current project phase
 3 Projects     tblABCDEFGHIJKL Owner        fldCCCCCCCCCCCCCC multipleRecordLinks Link to People table
 4 Projects     tblABCDEFGHIJKL Budget       fldDDDDDDDDDDDDDD currency           Approved budget (USD)
 ...
```

This flat format makes it easy to edit descriptions in bulk — open in a
spreadsheet, modify the `description` or `field_name` columns, and push the
changes back.

### Pushing changes with `air_meta_push()`

`air_meta_push()` accepts the tibble returned by `air_meta()` (possibly
modified), compares it against the current schema, and applies only the
changed rows via API PATCH requests:

```{r air-meta-push}
# Read the current metadata
meta <- air_meta("appXXXXXX")

# Edit descriptions for undocumented fields
meta$description[is.na(meta$description) & meta$field_name == "Start Date"] <-
  "ISO 8601 date when the project officially kicks off"

meta$description[is.na(meta$description) & meta$field_name == "End Date"] <-
  "Planned completion date; may be NA for open-ended projects"

# Push only the changed rows back to Airtable
air_meta_push("appXXXXXX", meta)
#> i Pushed 2 field changes.
```

`air_meta_push()` compares each row by `field_id` — so renaming a field in
the `field_name` column and pushing will rename it in Airtable as well. Type
changes are silently ignored because the Airtable API does not allow most
field type conversions.

### Making a base self-documenting with `air_meta_sync()`

`air_meta_sync()` takes documentation one step further: it writes the full
metadata tibble into a designated table *within the same base*. This means
anyone browsing the base in the Airtable UI can see a `_metadata` table
listing every table, field, type, and description — no external tooling
required.

```{r air-meta-sync}
# Upsert the current schema into a "_metadata" table in the same base
air_meta_sync("appXXXXXX")

# Use a custom table name
air_meta_sync("appXXXXXX", meta_table = "Data Dictionary")
```

The merge key is a composite of `table_name` and `field_name`, so running
`air_meta_sync()` repeatedly is idempotent — it updates existing rows and adds
new ones as the schema evolves.

**Workflow: documentation as code**

A typical pattern is to maintain `meta` as a CSV in your project repository,
edit it locally, and push changes in a single step:

```{r air-meta-workflow}
# On first run: export the current metadata to a file
meta <- air_meta("appXXXXXX")
write.csv(meta, "base-metadata.csv", row.names = FALSE, na = "")

# Later: edit base-metadata.csv in a text editor or spreadsheet, then:
meta <- read.csv("base-metadata.csv", na.strings = "")
air_meta_push("appXXXXXX", meta)
```

---

## Full backup with `air_dump()`

`air_dump()` creates a complete local snapshot of a base: the full schema
(as `schema.json`) plus all record data, one file per table. It is designed
to be a safe starting point for disaster recovery or base migration.

```{r air-dump-basic}
# Full backup including attachment files
air_dump("appXXXXXX", dir = "backup/my-base/")

# Backup metadata only for attachments (faster; skips downloading files)
air_dump("appXXXXXX", dir = "backup/my-base/", attachments = "meta")

# CSV format (flattens complex columns; good for spreadsheet portability)
air_dump("appXXXXXX", dir = "backup/my-base/", format = "csv")

# In-memory list (no files written; useful for inspection or small bases)
dump_list <- air_dump("appXXXXXX", format = "list")
names(dump_list)
#> [1] "schema"   "Projects" "People"   "Status Codes"
```

### Parameters

| Argument | Default | Notes |
|---|---|---|
| `base_id` | required | Base ID (starts with `app`) |
| `dir` | `NULL` (temp dir for json) | Output directory |
| `format` | `"list"` | `"list"`, `"json"`, or `"csv"` |
| `attachments` | `"file"` | `"file"` downloads; `"meta"` keeps URLs only; `"blob"` embeds in memory |
| `.token` | env var | Override the API token |

### Output structure (JSON format)

```
backup/my-base/
├── schema.json          # Full table/field schema
├── projects.json        # All records from the Projects table
├── people.json          # All records from the People table
├── status_codes.json    # All records from the Status Codes table
└── attachments/         # Downloaded attachment files (if attachments = "file")
    └── projects/
        └── recXXXXXXXX/
            └── proposal.pdf
```

Table names are sanitised for use as filenames: spaces and special characters
are replaced with underscores, and names are lowercased. `schema.json`
preserves the original names.

### What is included

- All records and all writable field values
- The full schema including field types and options (select choices, formula
  strings, currency symbols, etc.)
- Attachment metadata (URLs, filenames, MIME types) always; file contents when
  `attachments = "file"`

### What is NOT included

Computed and read-only fields are **excluded from data files** because they
cannot be written back via the API:

| Field type | Reason excluded from data |
|---|---|
| `formula` | Computed from other fields; formula string is in schema |
| `rollup` | Computed across linked records |
| `lookup` | Looks up values from a linked table |
| `count` | Counts linked records |
| `autoNumber` | Assigned by Airtable; cannot be set on write |
| `createdTime` | Set by Airtable on record creation |
| `lastModifiedTime` | Set by Airtable on last edit |
| `createdBy` | Set by Airtable |
| `lastModifiedBy` | Set by Airtable |
| `externalSyncSource` | Synced from external source |
| `aiText` | AI-generated |
| `button` | UI-only |

The schema captures these field definitions, so you can re-create them manually
in the web UI after a restore.

**Linked-record IDs are preserved** in the dump: `multipleRecordLinks` columns
contain the `recXXX` IDs of the linked records. However, after a restore, new
records receive new IDs, which means these links will be broken in the restored
base. See the [Re-linking after restore](#re-linking-after-restore-advanced)
section for the workaround.

---

## Restoring with `air_restore()`

`air_restore()` recreates a base from the output of `air_dump()`. It creates a
new base, adds all restorable fields to each table, and inserts the saved
records.

```{r air-restore-basic}
# Restore from a directory dump
new_base_id <- air_restore(
  "backup/my-base/",
  base_name = "My Base (restored 2025-06-04)",
  workspace_id = "wspXXXXXX"
)
#> i Creating base "My Base (restored 2025-06-04)"...
#> i Adding fields...
#> ! Field "Progress" (type formula) cannot be restored via the API - create it manually in the web UI.
#> ! Field "Record ID" (type autoNumber) cannot be restored via the API - create it manually in the web UI.
#> i Inserting records...
#> i Restore complete. New base ID: appYYYYYY.

# Restore from an in-memory list dump
new_base_id <- air_restore(
  dump_list,
  base_name = "My Base (copy)",
  workspace_id = "wspXXXXXX"
)
```

### Parameters

| Argument | Default | Notes |
|---|---|---|
| `dump` | required | List from `air_dump(format = "list")` or path to a dump directory |
| `base_name` | auto-generated | Name for the new base |
| `workspace_id` | env var | Workspace to create the base in |
| `attachments` | `"file"` | `"file"` re-uploads attachment files; `"meta"` skips upload |
| `attachment_dir` | inferred from `dump` | Override path to saved attachment files |
| `.token` | env var | Override the API token |

### What works

- All writable field types: text, numbers, dates, checkboxes, selects, URLs,
  emails, ratings, currencies, durations
- Formula fields: the formula string is translated from field IDs back to field
  names and re-created, though Airtable will mark it as `isValid: false` until
  all referenced fields exist
- Select choices: restored without their original choice IDs (Airtable assigns
  new IDs)
- Record data for all restorable fields

### Restore limitations

**Computed fields cannot be created via the API.** The following types are
silently skipped during field creation, and a warning is emitted for each:

- `formula`, `rollup`, `lookup`, `count`
- `autoNumber`, `createdTime`, `lastModifiedTime`, `createdBy`, `lastModifiedBy`
- `externalSyncSource`, `aiText`, `button`
- `multipleRecordLinks` (see below)

After a restore, open the base in the Airtable UI and manually re-create these
fields using the saved schema as a reference.

**Linked-record fields break on restore.** `multipleRecordLinks` fields store
lists of record IDs (e.g. `["recABC", "recDEF"]`). When records are inserted
into the restored base they receive *new* record IDs. The old IDs stored in
linked fields no longer refer to anything in the new base. For a workaround,
see [Re-linking after restore](#re-linking-after-restore-advanced).

**Record metadata cannot be preserved.** `airtable_id` (the original record
ID) and `airtable_created_time` are stripped before inserting records into the
new base. Records will have new IDs and new creation timestamps.

---

## Backup/restore workflow example

The following end-to-end example covers a typical backup-and-migrate scenario.
All chunks use `eval = FALSE` as they require live API credentials.

```{r workflow-dump}
library(airtable2)

# Step 1: Dump the source base to a local directory
air_dump(
  "appSOURCEBASE",
  dir     = "backup/",
  format  = "json",
  attachments = "file"   # download all attachment files
)
```

```{r workflow-inspect}
# Step 2: Inspect the schema to identify computed fields
schema <- jsonlite::read_json("backup/schema.json")

# Or use air_schema() on the live base
schema_tbl <- air_schema("appSOURCEBASE")
computed_types <- c(
  "formula", "rollup", "lookup", "count", "autoNumber",
  "createdTime", "lastModifiedTime", "createdBy", "lastModifiedBy"
)
needs_manual <- lapply(schema_tbl$fields, function(f) {
  f[f$type %in% computed_types, c("name", "type")]
})
names(needs_manual) <- schema_tbl$table_name
needs_manual
```

```{r workflow-restore}
# Step 3: Restore to a new base in the same (or different) workspace
new_id <- air_restore(
  "backup/",
  base_name    = "Projects (migrated)",
  workspace_id = "wspTARGETWORKSPACE"
)
```

```{r workflow-verify}
# Step 4: Verify record counts match
original <- air_schema("appSOURCEBASE")
restored  <- air_schema(new_id)

# Check table names
setdiff(original$table_name, restored$table_name)  # Should be character(0)
```

```{r workflow-manual}
# Step 5 (manual): In the Airtable UI, re-create computed fields
# (formulas, rollups, lookups, autoNumber, etc.) in the new base.
# Use the saved schema.json as a reference.

# Step 6: Re-link records if any tables use multipleRecordLinks
# (see the next section)
```

---

## Re-linking after restore (advanced)

When a base is restored, every record receives a new Airtable record ID. Any
`multipleRecordLinks` fields that were saved in the dump contain the *old* IDs,
which no longer point to valid records in the restored base. After a restore,
linked-record columns will be empty or invalid.

The general approach to repair links is:

1. Identify a **natural key** — a field in the linked table whose values are
   unique and stable across bases (e.g. a project code, an email address, or
   an explicit external ID field).
2. Build a mapping from old record IDs to new record IDs using that key.
3. Rewrite the linked-record columns in the restored base using the new IDs.

```{r relinking-example}
# Illustrative approach — adapt to your specific tables and keys

# Read the new IDs from the restored base (People table has email as natural key)
people_new <- air_read(new_id, "People")
# people_new has new `airtable_id` values, same `Email` values

# Read the old data from the dump (load the JSON or use the in-memory dump)
dump <- air_dump("appSOURCEBASE", format = "list")
people_old <- dump[["People"]]

# Build a mapping: old_id -> new_id via shared Email column
id_map <- merge(
  people_old[, c("airtable_id", "Email")],
  people_new[, c("airtable_id", "Email")],
  by = "Email",
  suffixes = c("_old", "_new")
)

# Rewrite the Owner column in the Projects table
projects_new <- air_read(new_id, "Projects")

remap_links <- function(old_ids, id_map) {
  # old_ids is a list-column element (character vector of record IDs)
  if (is.null(old_ids) || length(old_ids) == 0) return(old_ids)
  new_ids <- id_map$airtable_id_new[match(old_ids, id_map$airtable_id_old)]
  new_ids[!is.na(new_ids)]
}

projects_new$Owner <- lapply(projects_new$Owner, remap_links, id_map = id_map)

# Write the corrected links back
air_upsert(
  projects_new[, c("airtable_id", "Owner")],
  base_id  = new_id,
  table    = "Projects",
  merge_on = "airtable_id"
)
```

A planned `restore_linked_fields()` helper will automate this pattern. Track
progress and add your use case at the
[GitHub issues page](https://github.com/One-Health-Research-Consulting/airtabler/issues).
